### 2018-07-23 contents/abstract.tex
final sync, 2-sided print correction

\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches. TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear. Single-tensor diffusion tensor neuroimaging (DTI) studies of CN V revealed diffusivity changes in the cistern segment. However, the portion of the nerve within the brainstem remained elusive due to the limits of DTI, and the lack of reliable software to researchers. Thereby the study of TN using diffusion imaging is time-consuming and error-prone due to the complete manual involvement in all steps of data process, placement of region-of-interest (ROI), and tractography parameter tuning, and data analysis. 
There are two aims of this thesis: 1) to develop a fully-automated software framework to collate and analyze diffusion tractography to reduce human error and increase iteration speed of experiments. 2) to apply the method developed to the analysis of the trigeminal system in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further. 
The specific aims are: a) establish the feasibility of high angular resolution diffusion tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error allow non-linear registration of tractography across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway. 
Towards these goals, in Study I, we establish the feasibility of applying spherical deconvolution tractography to delineate the full course of CNV from the trigeminal nucleus and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN can be distinguished by disruptions in the brainstem course of the nerve. In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image. In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) processing pipeline framework. Finally, in Study IV, we deploy end-to-end machine-learning TN classification to automatically discover diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway. 
In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification. The application of this system to TN revealed specific diffusivity changes in all three segments of the trigeminal CN V, pontine decussation, and S1  white matter pathways, and pin-points the exact location of the diffusivity disruption at millimetre level.
\end{abstract}

>>>>

\begin{abstract}
% \addcontentsline{toc}{chapter}{Abstract}
Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches. TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear. Single-tensor diffusion tensor neuroimaging (DTI) studies of CNV revealed diffusivity changes in the cistern segment. However, the portion of the nerve within the brainstem remained elusive due to DTI limits. Diffusion imaging of TN is also error-prone due to manual data processing and analysis.

This thesis aims to: 1) develop a fully-automated software framework to analyze diffusion tractography to reduce error and increase speed of experiments. 2) to apply the method developed to the analysis of the trigeminal system in a more substantial patient group, and apply state-of-the-art methods to advance the study of TN further.

The specific aims are: a) establish the feasibility of multi-tensor tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway. 
Towards these goals, in Study I, we establish the feasibility of applying multi-tensor tractography to delineate the full course of CNV and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN by disruptions in the brainstem course of the nerve. In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image. In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) processing pipeline framework. Finally, in Study IV, we deploy end-to-end machine-learning TN classification to automatically discover diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway.

In sum, this thesis presents a detailed road-map of the development and application of end-to-end diffusion tractography machine learning classification. The application to TN revealed specific diffusivity changes in trigeminal CN V, pontine, and S1 white matter pathways, and pin-points the locations of the diffusivity disruptions at millimetre level.
\end{abstract}

### 2018-07-23 contents/contributions.tex
final sync, 2-sided print correction

\chapter*{Statement of Contributions}
\addcontentsline{toc}{chapter}{Statement of Contributions}

@@

\noindent
Dr. Mojgan Hodaie (supervisor) - mentorship, assistance with experiment design, assistance with data interpretation, and assistance with thes
is preparation.

@@

\noindent
Dr. Jidan Zhong - assistance with experiment design, assistance with data interpretation, assitance with method development

>>>>

\chapter{Statement of Contributions}
% \addcontentsline{toc}{chapter}{Statement of Contributions}

@@

\noindent
Dr. Mojgan Hodaie (supervisor) - mentorship, assistance with experiment design, assistance with data interpretation, and assistance with thesis preparation.

@@

\noindent
Dr. Jidan Zhong - assistance with experiment design, assistance with data interpretation, assistance with method development

### 2018-07-23 contents/hypotheses.tex
final sync, 2-sided print correction

\section{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibres in the pons. However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals. Previous single-tensor tractography studies of CN V have been limited to the imaging of the cisternal nerve segment. In this study, we use multi-tensor tractography with two aims: first, we focus on the delineation of the brainstem sub-segments of CN V; and the second, on the feasibility of applying multi-tensor tractography in the study of diffusivity differences in TN and MS-TN.

@@

\section{Study II: Diffusion Weighted Image to T1 Co-registration Without Reversed-blip: Investigation of Best Practices}
The task of establishing inter-subject anatomical correspondences is imperative for extending tractography to multiple subjects. Population studies with inter-subject T1 registration are well-established, where a normalized T1 template is generated from numerous subjects to facilitate data analysis. Similarly, DWI of each subject can be co-registered to their corresponding T1, to extend the T1 template space to tractography space. Therefore the DWI-T1 co-registration accuracy is crucial to the reliability of inter-subject tractography alignment. DWI-T1 co-registration is challenging due to the DWI eddy-current and EPI field distortions. While new DWI sequences with reversed-blips have been shown to reduce non-linear distortions, they have yet to be commonly adopted in clinical practice. While T1 inter-subject registration methodology is well documented, the feasibility, and strategy for DWI-T1 co-registration on sequence without reversed-blips have not been documented, and forms a potential bottleneck on the performance of any DWI-T1 population registration pipeline.

@@

\section{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
The construction of an automated group tractography software framework depends on CN V brainstem delineation using multi-tensor tractography and the optimal DWI-T1 co-registration strategy. The aim is to create and validate a software framework that automates the tractography process on a large-scale, and also streamline research workflow by providing region-of-interest, and tractography parameter management. The software also provides the ability to report the tractography output at a group level both visually, as well as quantifying the consistency of group tractography output by developing a rating score called Normalized Overlap Score (NOS) score.

@@

\section{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
The methodology developed in the previous study was applied to the study of the trigeminal sensory pathway. The aim of this study was to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group. The pathway was divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment. Along-the-tract analysis were carried out by dividing each segment into 30 subdivisions. Following this, we used machine learning algorithms to auto-determine relevant subdivisions that maximally differentiates TN from controls.

>>>>

\section[Study I]{Study I: Diffusivity signatures characterize trigeminal neuralgia associated with multiple sclerosis}
MS-TN is thought to relate to MS lesions in the brainstem, in the area of the trigeminal fibres in the pons. However the exact anatomical relationship between the lesions and trigeminal nerve within the brainstem has been difficult to quantify, due to the limited contrast and resolution in standard MRI and the variability of MS lesion across individuals. Previous single-tensor tractography studies of CN V have been limited to the imaging of the cisternal nerve segment. In this study, we use multi-tensor tractography with two aims: first, we focus on the delineation of the brainstem sub-segments of CN V; and the second, on the feasibility of applying multi-tensor tractography in the study of diffusivity differences in TN and MS-TN.

@@

\section[Study II]{Study II: Diffusion Weighted Image to T1 Co-registration Without Reversed-blip: Investigation of Best Practices}
The task of establishing inter-subject anatomical correspondences is imperative for extending tractography to multiple subjects. Population studies with inter-subject T1 registration are well-established, where a normalized T1 template is generated from numerous subjects to facilitate data analysis. Similarly, DWI of each subject can be co-registered to their corresponding T1, to extend the T1 template space to tractography space. Therefore the DWI-T1 co-registration accuracy is crucial to the reliability of inter-subject tractography alignment. DWI-T1 co-registration is challenging due to the DWI eddy-current and EPI field distortions. While new DWI sequences with reversed-blips have been shown to reduce non-linear distortions, they have yet to be commonly adopted in clinical practice. While T1 inter-subject registration methodology is well documented, the feasibility, and strategy for DWI-T1 co-registration on sequence without reversed-blips have not been documented, and forms a potential bottleneck on the performance of any DWI-T1 population registration pipeline.

@@

\section[Study III]{Study III: Merged Group Tractography Evaluation with Selective Automated Group Integrated Tractography}
The construction of an automated group tractography software framework depends on CN V brainstem delineation using multi-tensor tractography and the optimal DWI-T1 co-registration strategy. The aim is to create and validate a software framework that automates the tractography process on a large-scale, and also streamline research workflow by providing region-of-interest, and tractography parameter management. The software also provides the ability to report the tractography output at a group level both visually, as well as quantifying the consistency of group tractography output by developing a rating score called Normalized Overlap Score (NOS) score.

@@

\section[Study IV]{Study IV: Gaussian Process Classification of Trigeminal Neuralgia With Merged Tractography}
The methodology developed in the previous study was applied to the study of the trigeminal sensory pathway. The aim of this study was to examine the diffusivity signature differences in the trigeminal sensory pathway between TN and controls in a larger group. The pathway was divided into three major regions: CN V, trigeminopontothalamic (TPT) segment, and the thalamocortical S1 segment. Along-the-tract analysis were carried out by dividing each segment into 30 subdivisions. Following this, we used machine learning algorithms to auto-determine relevant subdivisions that maximally differentiates TN from controls.

### 2018-07-23 contents/paper1.tex
figure layout fixes
final sync, 2-sided print correction

Understanding the microstructural anatomy of the trigeminal fibres is essential to determine how MS affects CN V brainstem fibres and how alterations to CN V in MS are related to the clinical presentation of pain. Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure 1A-B) \cite{Cruccu2009,DaSilva2005}. Conventional MR imaging cannot visualize in isolation the brainstem fibres of CN V. Furthermore conventional tractography cannot resolve the brainstem crossing fibres \cite{Farquharson2013}, which leads to unreliable CN V delineations within the brainstem, and inaccuracies in diffusivity measurements.

@@

\section{Results}
\subsection{MTT permits visualization of brainstem trigeminal fibres}
MTT allowed for the visualization of brainstem CN V fibres in isolation, by distinguishing them from the surrounding crossing cerebellar peduncular fibres. The SDT method could not adequately delineate the nerve (Figure 2). Comparison of tractography results between SDT and MTT showed that SDT was unable to differentiate fibres crossing the cerebellar peduncle fibres, while MTT delineated the brainstem course of CN V into the region of the trigeminal main sensory nucleus. Delineation of MS-TN CN V by MTT revealed that the tracts became visibly sparse as they came into proximity of the MS plaque (Figure 3A). FA decreases in proximity to the MS plaque could also be visualized by colour shift in the tractography model with more orange/yellow colour presented. The fibres of the unaffected CN V did not exhibit this change, and extended caudally into the trigeminal nuclei (Figure 3B).

@@

\subsubsection{Cisternal changes are unique in TN}
In the cistern segment (Figure 4), statistically significant differences are found only between the TN group and other groups. No significant differences were found in between segments of other subject groups. Specifically, the FA measure of symptomatic side (+) in TN patients is shown to be significantly greater compared to all of the other patient groups and segments, while the asymptomatic side (-) in TN patients does not show any significant differences compared to other groups. In RD, AD and MD measurements, the only significant differences are observed between TN(+) and TN(-), where TN(+) RD/AD/MD are lower than that of TN(-). No symptomatic/asymptomatic (+/-) or left/right differences were observed in MS-TN and control groups. These results suggest that the cisternal CN V diffusivity changes are unique to the TN group.

\subsubsection{REZ changes are found predominantly in TN}
In the REZ (Figure 5), statistically significant results are found within, but not unique to TN. TN(+) FA is shown to be significantly lower than TN(-) FA, while TN(+) RD and MD are shown to be significantly greater than TN(-); TN(+) MD and RD are shown to be significantly greater compared to all other groups and sides; while TN AD showed no significant differences. In contrast, MS-TN patients showed no significant differences between MS-TN(+) and MS-TN(-) sides in any of the diffusivity metrics, suggesting no intra-group differences. Comparisons between MS-TN(+/-) and TN(+/-) revealed no significant inter-group FA differences. However, both TN and MS-TN groups are shown to be different from controls. Both TN(+/-) and MS-TN(+/-) REZ have significantly lower FA than that of control REZ; control AD and RD are shown to be significantly lower than TN(+/-); while there is no significant difference between left and right sides in controls. 
\subsubsection{Pontine diffusivities showed no significant changes in all groups}
In the pontine tissue (Figure 6), there are no significant intra-group or inter-group differences in FA, RD and MD for all groups and sides. However MS-TN(-) AD is significantly higher than TN(+/-) and controls. No intra-group AD differences were found in MS-TN.
2D. Diffusivities in peri-lesional trigeminal fibres highlight MS-TN pain

In the peri-lesional region (Figure 7), the patterns shows significant intra-group and inter-group differences in MS-TN. MS-TN(+) FA is significantly lower than that of MS-TN(-), as well as significantly lower than TN(+/-)and Controls, while MS-TN(-) FA showed no inter-group differences with TN and Controls. RD, and MD showed only intra-group differences between MS-TN(+) and MS-TN(-),where MS-TN(+) RD and MD are significantly higher than that of MS-TN(-). AD showed no significant differences. The pattern of differences for peri-lesional region is unique to MS-TN.

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure5.png}
\caption{In the REZ, MS-TN, and TN patients have significantly different diffusivities when compared to controls. TN+ also shows significantly different FA, RD, and MD from TN-. Moreover TN+ shows significant different from MS-TN+ as well. Please refer to Figure 4 for explanations of layout and annotations.}
\centering
\label{fig:MSfigure5}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure6.png}
\caption{In the pontine segment unaffected by MS plaques, there are little differences between groups. However MS-TN- shows significantly different AD than TN- and controls. Please refer to Figure 4 for explanations of layout and annotations.}
\centering
\label{fig:MSfigure6}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure7.png}
\caption{The peri-lesional segment highlighting MS-TN+ specific diffusivity differences. Both MS-TN+ and MS-TN- show significantly lowered FA than all the other groups. RD of MS-TN- is also significantly lower than that of MS-TN+. Please refer to Figure 4 for explanations of layout and annotations.
http://msj.sagepu}
\centering
\label{fig:MSfigure7}
\end{figure}

@@

We demonstrate for the first time that diffusion MRI tractography is capable of anatomically distinguishing between TN and MS-TN. Analyses revealed unique, focal MS-TN diffusivity changes along CN V. These changes highlight the differences in diffusivities between MS-TN, TN and healthy controls, thereby permitting their use as neuroimaging signatures for distinguishing these conditions. 
Previous studies of the CN V with DTI tractography have been met with limitations of the single-tensor Gaussian diffusion model, and were incapable of resolving brainstem trigeminal tracts (Figure 2A). Recent improvements in multi-tensor imaging are now capable of more complex diffusion profile of crossing fibres in neural tissue, thereby permitting reliable reconstruction of the brainstem CN V (Figure 2B), and subsequently the study of CN V changes in MS patients (Figure 3).

>>>>

Understanding the microstructural anatomy of the trigeminal fibres is essential to determine how MS affects CN V brainstem fibres and how alterations to CN V in MS are related to the clinical presentation of pain. Direct comparisons between TN and MS-TN at the level of CN V brainstem fibres has been challenging because it is difficult to reliably identify, segment, and measure trigeminal anatomy with conventional magnetic resonance imaging (MRI)(Figure \ref{fig:MSfigure1}A-B) \cite{Cruccu2009,DaSilva2005}. Conventional MR imaging cannot visualize in isolation the brainstem fibres of CN V. Furthermore conventional tractography cannot resolve the brainstem crossing fibres \cite{Farquharson2013}, which leads to unreliable CN V delineations within the brainstem, and inaccuracies in diffusivity measurements.

@@

\section{Results}
\subsection{MTT permits visualization of brainstem trigeminal fibres}
MTT allowed for the visualization of brainstem CN V fibres in isolation, by distinguishing them from the surrounding crossing cerebellar peduncular fibres. The SDT method could not adequately delineate the nerve (Figure \ref{fig:MSfigure2}). Comparison of tractography results between SDT and MTT showed that SDT was unable to differentiate fibres crossing the cerebellar peduncle fibres, while MTT delineated the brainstem course of CN V into the region of the trigeminal main sensory nucleus. Delineation of MS-TN CN V by MTT revealed that the tracts became visibly sparse as they came into proximity of the MS plaque (Figure \ref{fig:MSfigure3}A). FA decreases in proximity to the MS plaque could also be visualized by colour shift in the tractography model with more orange/yellow colour presented. The fibres of the unaffected CN V did not exhibit this change, and extended caudally into the trigeminal nuclei (Figure \ref{fig:MSfigure3}B).

@@

\subsubsection{Cisternal changes are unique in TN}
In the cistern segment (Figure \ref{fig:MSfigure4}), statistically significant differences are found only between the TN group and other groups. No significant differences were found in between segments of other subject groups. Specifically, the FA measure of symptomatic side (+) in TN patients is shown to be significantly greater compared to all of the other patient groups and segments, while the asymptomatic side (-) in TN patients does not show any significant differences compared to other groups. In RD, AD and MD measurements, the only significant differences are observed between TN(+) and TN(-), where TN(+) RD/AD/MD are lower than that of TN(-). No symptomatic/asymptomatic (+/-) or left/right differences were observed in MS-TN and control groups. These results suggest that the cisternal CN V diffusivity changes are unique to the TN group.

\subsubsection{REZ changes are found predominantly in TN}
In the REZ (Figure \ref{fig:MSfigure5}), statistically significant results are found within, but not unique to TN. TN(+) FA is shown to be significantly lower than TN(-) FA, while TN(+) RD and MD are shown to be significantly greater than TN(-); TN(+) MD and RD are shown to be significantly greater compared to all other groups and sides; while TN AD showed no significant differences. In contrast, MS-TN patients showed no significant differences between MS-TN(+) and MS-TN(-) sides in any of the diffusivity metrics, suggesting no intra-group differences. Comparisons between MS-TN(+/-) and TN(+/-) revealed no significant inter-group FA differences. However, both TN and MS-TN groups are shown to be different from controls. Both TN(+/-) and MS-TN(+/-) REZ have significantly lower FA than that of control REZ; control AD and RD are shown to be significantly lower than TN(+/-); while there is no significant difference between left and right sides in controls. 
\subsubsection{Pontine diffusivities showed no significant changes in all groups}
In the pontine tissue (Figure \ref{fig:MSfigure6}), there are no significant intra-group or inter-group differences in FA, RD and MD for all groups and sides. However MS-TN(-) AD is significantly higher than TN(+/-) and controls. No intra-group AD differences were found in MS-TN.
2D. Diffusivities in peri-lesional trigeminal fibres highlight MS-TN pain

In the peri-lesional region (Figure \ref{fig:MSfigure7}), the patterns shows significant intra-group and inter-group differences in MS-TN. MS-TN(+) FA is significantly lower than that of MS-TN(-), as well as significantly lower than TN(+/-)and Controls, while MS-TN(-) FA showed no inter-group differences with TN and Controls. RD, and MD showed only intra-group differences between MS-TN(+) and MS-TN(-),where MS-TN(+) RD and MD are significantly higher than that of MS-TN(-). AD showed no significant differences. The pattern of differences for peri-lesional region is unique to MS-TN.

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure5.png}
\caption{In the REZ, MS-TN, and TN patients have significantly different diffusivities when compared to controls. TN+ also shows significantly different FA, RD, and MD from TN-. Moreover TN+ shows significant different from MS-TN+ as well. Please refer to Figure \protect\ref{fig:MSfigure4} for explanations of layout and annotations.}
\centering
\label{fig:MSfigure5}
\end{figure}

\begin{figure}[p]
\includegraphics[width=\textwidth]{figure6.png}
\caption{In the pontine segment unaffected by MS plaques, there are little differences between groups. However MS-TN- shows significantly different AD than TN- and controls. Please refer to Figure \protect\ref{fig:MSfigure4} for explanations of layout and annotations.}
\centering
\label{fig:MSfigure6}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{figure7.png}
\caption{The peri-lesional segment highlighting MS-TN+ specific diffusivity differences. Both MS-TN+ and MS-TN- show significantly lowered FA than all the other groups. RD of MS-TN- is also significantly lower than that of MS-TN+. Please refer to Figure \protect\ref{fig:MSfigure4} for explanations of layout and annotations.}
\centering
\label{fig:MSfigure7}
\end{figure}

@@

We demonstrate for the first time that diffusion MRI tractography is capable of anatomically distinguishing between TN and MS-TN. Analyses revealed unique, focal MS-TN diffusivity changes along CN V. These changes highlight the differences in diffusivities between MS-TN, TN and healthy controls, thereby permitting their use as neuroimaging signatures for distinguishing these conditions. 
Previous studies of the CN V with DTI tractography have been met with limitations of the single-tensor Gaussian diffusion model, and were incapable of resolving brainstem trigeminal tracts (Figure \ref{fig:MSfigure2}A). Recent improvements in multi-tensor imaging are now capable of more complex diffusion profile of crossing fibres in neural tissue, thereby permitting reliable reconstruction of the brainstem CN V (Figure \ref{fig:MSfigure2}B), and subsequently the study of CN V changes in MS patients (Figure \ref{fig:MSfigure3}).
